---
title: "R Notebook"
output: html_notebook
---

# Load required packages
```{r}
library(tidyverse)
library(here)
```

# Read in cleaned data
```{r}
clean_candy <- read_csv(here("clean_data/cleaned_candy_data"))

```

# Question 1 - What is the total number of candy ratings given across the three years. (number of candy ratings, not number of raters. Don’t count missing values)
```{r}
clean_candy %>% 
  filter(!is.na(candy_feelings)) %>% 
  nrow()

```

# Question 2 - What was the average age of people who are going out trick or treating and the average age of people 3. not going trick or treating?
```{r}
clean_candy %>% 
  filter(trick_or_treating == "Yes") %>% 
  summarise(mean(age))
```

```{r}
clean_candy %>% 
  filter(trick_or_treating == "No") %>% 
  summarise(mean(age))
```

# Question 3 - For each of joy, despair and meh, which candy bar revived the most of these ratings?
```{r}
clean_candy %>% 
  filter(candy_feelings == "joy") %>% 
  summarise(mean(age))
```















