---
title: "seabirds_report"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to Dataset

The seabirds data set consisted of an .xls file, which contained two main 
worksheets.  The other two worksheets contained a key of abbreviations to assist
in understanding what the variables were in each worksheet.

The data records counts of seabirds of different species observed from ships 
during the period 31st July 1969 to 21st December 1990.  Location of recordings
(by latitude and longitude) where also recorded in the dataset.  

## List of Assumptions

In looking at the analysis required, it was assumed that information on specific
dates and times would not be required and it was decided that cleaning was not
really required on these particular columns.

It was also assumed that many of the columns in the raw data would not be required
and these would also be dropped at the initial stages of cleaning.

Assumption was also taken that all the bird count figures were accurate.  The 
range was established as count of 1 up to 99,999(6 observations).  It is possible 
that the latter figures are erroneous, however some research confirmed that 
colonies of certain birds (like the short tailed shearwater) can be upwards of 
millions so this data had to be taken at face value.

## Cleaning the Data

1. Bird and ship worksheets were read in as separate tables using read_xls.
2. Column names were cleaned using the _janitor_ package clean_names, with some 
requiring re-naming due to excessive name length.
3. Columns required for analysis (including all names, latitude/longitude) were
  selected.
4. A __LEFT__ join was then performed on the birds and ship data to compile the
data into one table.
5. Two hard-coded strings were then removed from the _time_ and 
_species_common_name_ columns as shown below:

```{r, eval=FALSE}
#--------------Recoding time column--------------------------------------------

bird_observations <- bird_observations %>% 
  mutate(time = str_remove(time, pattern = "1899-12-31 "))

# -------Removing the rows where no bird counts were definitely recorded-------

bird_observations <- bird_observations %>% 
  filter(!str_detect(species_common_name, pattern = "NO BIRDS RECORDED"))

```

6. The _species_common_name_ then required a number of cleaning steps to remove
additional variables that had been recorded in the same string as the species
name (such as bird age category or plummage colours).  Regex patterns were used
to remove these erroneous strings, as it was noted that they consisted of 2 or 
more capital letters in a row, or capitals followed by numbers.  It was also 
observed that there was some additional white space in some of the observations
so this was removed with the _trim_ws_ function:

```{r, eval=FALSE}
bird_observations <- bird_observations %>% 
  mutate(species_common_name = 
           str_remove(species_common_name, 
                      pattern = ("[A-Z]{2,}[0-9]*")
           )) %>%
  mutate(species_common_name =
           str_remove(species_common_name,
                      pattern = ("[A-Z]{2,}")
           )) %>% 
  
  mutate(species_common_name = 
           str_remove(species_common_name, 
                      pattern = " sensu lato"
           )) %>% 
  
  mutate(species_common_name = 
           str_remove(species_common_name, 
                      pattern = "[(][A-z]+[)]"
           )) %>% 
  
  mutate(species_common_name = 
           str_remove(species_common_name, 
                      pattern = "[ ]M$"
           )) %>% 
  
  mutate(species_common_name =
           trimws(species_common_name))
```

7. Similar cleaning steps were performed on _species_scientific_name_ and 
_species_abbreviation_ , with slight differences in the regex patterns used.

8. It was decided to recode the _month_ column, in case it would become useful 
in future analyses:

```{r, eval=FALSE}
bird_observations <- bird_observations %>% 
  mutate(month = recode(month, 
                        "1" = "January",
                        "2" = "February",
                        "3" = "March",
                        "4" = "April",
                        "5" = "May",
                        "6" = "June",
                        "7" = "July",
                        "8" = "August",
                        "9" = "September",
                        "10" = "October",
                        "11" = "November",
                        "12" = "December")
  )
```

9. There were a number of observations where the count of birds were recorded as 
N/A.  It was decided not to drop these as there had been similar N/A values 
removed where _species_common_name_ had been recorded as "NO BIRDS RECORDED".
It was therefore decided to impute these values as a one-off sighting and give 
them a count value of 1:

```{r, eval=FALSE}
bird_observations <- bird_observations %>% 
  mutate(bird_count = coalesce(bird_count, 1, na.rm = TRUE))
```

10. The final cleaning step required the data be written to clean_data and this 
was achieved using the _write_csv_ function:

```{r, eval=FALSE}
write_csv(bird_observations, file = "clean_data/cleaned_bird_observations")
```


## Results and Analysis

__Q1. Which bird had the most individual sightings?__  
      Wandering albatross  
      Diomedea antipodensis / exulans  
      DIOANTEXU  

__Q2. Which bird had the highest total count?__ 
      Short-tailed shearwater  	
      Puffinus tenuirostris  	
      PUFTEN  
    
__Q3. Which bird had the highest total count above a latitude of -30?__  
      Wedge-tailed shearwater  	
      Puffinus pacificus  	
      PUFPAC  
    
__Q4. How many different types of birds were only ever seen in groups of 1?__  
      64  

__Q5. How many penguins were seen?__  
      162  


## Additional Analysis

No additional analysis carried out at this time (I am running out of time to do 
the candy data set!)






















